Thermal reactionomes reveal divergent responses to thermal extremes in warm

### 2015-07-02 doc/mec-14-1243-response-letter.md

	To test whether the temperature at which thermally-responsive transcripts were activated differs between species, we identified the temperature at which there was the greatest change in expression for each transcript in each species, using only the transcripts with a significant species x temperature interaction.

> -207-216 –is t test appropriate here, given that data are categorical rather than continuous? (Expression is observed at 12 distinct temperatures).
>

In this case, we aren't testing expression across the 12 temperatures. For each transcript, we used the predicted expression values from the statistical model to identify the temperature where expression had the greatest change (the temperature of transcript activation). We then compared the vectors of temperature of activation (a continuous variable) between species using the t-test. We now explain this in the text (lns 297 - 299):

	We then performed a t-test to determine if the mean temperature of transcript activation differed between the two species for each group.

> -235-245 Something’s not adding up about the number of transcripts in each category 237-239 says, “Of the Formicidae hits, we identified 9,052 transcripts, or 8% of the total, with expression patterns that were thermally responsive” (implying that the 9052 thermally responsive transcripts all had hits to Formicidae) But then later in the paragraph, “The proportion of responsive transcripts is the same if we focus only on those transcript (4,618 transcripts out of 51,246).” (Implying that only half of the thermally responsive transcripts had blast hits). 
>

This was our mistake as the phrase “Of the Formicidae hits..." was left in unintentionally from a previous draft. The percentage of responsive transcripts is out of the total.

> 414-418 “In both species, there are 3 -- 4 times more transcripts upregulated at low than high temperatures (Table 1). A potential explanation for this pattern is that selection has been stronger on upper than lower thermal limits, thereby reducing genetic variation at high temperatures (Kingsolver et al. 2013; Kelly et al. 2013).” I would argue that the data show reduced plasticity of gene expression at higher temperatures rather than reduced genetic variation (= variation among individuals).
>

As this section raised concern from multiple reviewers and is not essential to our main conclusions, we have removed it from the manuscript.

> Reviewer: 3
> 
> Recommendation: Accept, minor revisions
> 
> Comments:
> 
> MEC-14-1243- MS TITLE: Thermal reactionomes reveal adaptive responses to thermal extremes in warm and cool-climate ant species
> 
> This paper brings valuable contribution to our understanding of thermal adaptation of two Ant species living in contrasted thermal regions. The authors found that a significant part of the transcriptome was thermally responsive. For both species, more transcripts responded to low than high temperatures, a result that I find counterintuitive and unexpected. The authors also found differential thermal-responsive transcriptional responses between the two species suggesting different thermal adaptation (and putative underlying mechanisms).
> 
> The authors present a set of experiments with insects that were assessed with a range of thermal conditions (12), ranging from lower to upper thermal limits, as well as transcriptional changes in reaction to temperature exposure (RNA seq). The topic is timely with the increasing interest in the molecular and physiological basis of response to temperature/climate change. The nature of cold response/tolerance is highly complex which makes the effort to unravel it very difficult.
> 
> Experiments were carefully conducted under well controlled conditions. The paper is logically arranged and well written, in good English. The methods used in this study were appropriate and the experiments were correctly conducted which makes the data set reliable. I also particularly appreciated that this is an hypothesis-driven study.
> 
> In spite of this, I have a number comments and queries that needs to be answered or at least discussed (see below).
> 
> -Abstract and elsewhere in MS : “enhanced defense hypothesis” . I find the term “defense” a bit confusing as it often used for immune process (defense genes …). “Stronger induced response hypothesis” could be an alternative.
>

We agree with this comment and have changed the ‘enhanced defense’ to the ‘enhanced response’ hypothesis throughout the paper.

> -L77: please provide factual examples for “shift in alternate suite of tolerance genes”

We have added citations to Neelakanta et al (2012), who showed that inducing expression of an antifreeze glycoprotein can increase the ability of fruit flies to tolerate cold, and Franssen et al. (2011), who showed that a southern population of Zosteria marina is more resilient to heat shock than a northern population (ln 131).

> 
> -L61-86: I am not totally convinced that the whole set of thermal-responsive genes will coordinately change in such way that it will fit to one of the three proposed mechanistic hypotheses of thermal adaptation (1 defense, 2 tolerance or 3 assimilation). I assume that transcriptional response will be more complex and gene-specific. Some genes will fit the first hypothesis, while other will fit the others. Please comment.
>

The reviewer is almost certainly correct, and it is our goal to determine the relative proportion of the transcriptome that falls into each category. We state this in the final sentence of the paragraph (lns 138-139:

Of course, these hypotheses are not mutually exclusive. By taking a reactionome approach, we can quantify if, and under what conditions, these mechanisms contribute to thermal adaptation.

> -L107: why such a small number of individuals (5) were used to monitor thermal limits?
>

A previous study (Warren and Chick 2013) used larger sample sizes (10 ants per colony) to demonstrate differences in the thermal limits of these species. Our measurement of thermal limits were simply a confirmation of their results.

> -L109-120: Ten ants were exposed to 12 different temperatures for each species. I may have misunderstood but it is not clear to me if there were replicates for each temperature?

Please see responses to reviewer 1 on this topic.

> 
> -L109-115: Twelve temperatures were used to establish a reaction norm of transcript expression patterns. Expression was checked at a single time point : after one hour of temperature exposure. It is well known that many thermal-responsive genes show temporal patterns of expression peaking at various time after exposure (see Sorensen 2005 Cell Stress & Chaperones  10, 312–328). Here, interpretations are based on early responsive genes (1h of exposure), which might only explain a small fraction of molecular response to temperature. Instead of taking 12 temperatures, it could have been of interest to check temporal patterns, perhaps with less thermal conditions. In fact, were 12 temperatures truly necessary to establish a reaction norm? My point is that this study should acknowledge that a large suite of thermally-responsive transcripts are overlooked and cannot be captured by the experimental design.

The temporal dynamics of expression response to thermal response are clearly crucial. We acknowledge this complexity by specifying that our reactionome is focused on capturing a wide temperature gradient, at the expense of having to focus only on a single time-point in the temporal response. Specifically, after defining the reactionome in the Intro we state (lns 94-97:

	Although temporal patterns of transcriptional activity (e.g. fast- vs. slow- responding genes) are also important components of an organism's transcriptional response to environmental conditions (Sørensen et al. 2005), we focus here on the response of transcripts across conditions at the same time point.

And in the Methods (ln 197):

	Thus, our reactionome characterized early, but not late, responding genes.

Finally, in the section on caveats in the Discussion (ln 554-556):

	… species may differ in gene expression along axes which we have not measured here, primarily temporal patterns of gene expression (Sorensen et al. 2005).

> 
> -Microarray and NGS data are regularly validated by qPCR. I am surprised that authors did not include validation step (on 5-10 genes) in their experiments. Previous studies have shown extremely close correlations between qPCR and RNAseq data [1-4]. Ideally, the authors should re-validate their findings (potentially by qPCR) in a separate cohort of samples.

Please see our general comments above.

> 
> 1.Griffith M, Griffith OL, Mwenifumbo J, Goya R, Morrissy a S, et al. (2010) Alternative expression analysis by RNA sequencing. Nat Methods 7: 843–847. doi:10.1038/nmeth.1503.
> 
> 2.Asmann YW, Klee EW, Thompson EA, Perez E a, Middha S, et al. (2009) 3’ tag digital gene expression profiling of human brain and universal reference RNA using Illumina Genome Analyzer. BMC Genomics 10: 531. doi:10.1186/1471-2164-10-531.
> 
> 3.Wu AR, Neff NF, Kalisky T, Dalerba P, Treutlein B, et al. (2014) Quantitative assessment of single-cell RNA-sequencing methods. Nat Methods 11: 41–46. doi:10.1038/nmeth.2694.
> 
> 4.Shi Y, He M (2014) Differential gene expression identified by RNA-Seq and qPCR in two sizes of pearl oyster (Pinctada fucata). Gene 538: 313–322. doi:10.1016/j.gene.2014.01.031.
> 
> -L197: comparative analyses between species were performed by focusing on transcripts that had species x temperature interaction. What about genes that had unique profile, that is, genes that were species-specific? These might contribute to thermal adaptation. How these genes were treated?

The species-specific genes were included in the analysis. In Table 1, these are the genes that were expressed in one species, but were “Not Responsive” or missing in the other species. We discuss some of these genes in the paragraph in the Discussion that begins “In contrast to cold tolerance, the enhanced thermal limit in A. picea is best explained by the tolerance hypothesis…” (ln 469 - 489).

> 
> -L196-201 & 207-216. It wasn’t clear to me why authors arbitrary defined 5 expression patterns (categories) ? Authors could have use k-means algorithm that can be used to partition the input data set into k partitions (clusters). In such way, main groups of temperature-responsive genes could be identified by K-mean clustering without a priori, and this would also provide statistical measures of over-representation (EASE scores).
>

The expression categories were not arbitrary, but represent a priori predictions that are biologically meaningful and can be related to the predictions of the three hypotheses (lns 272 - 275). While statistical clustering may be valuable, it would make hypothesis testing difficult. What would one do with a cluster that increased at 26 degrees, decreased at 29 and then leveled off, for example? This could be explored in another paper, but is not a good analytical fit with the goal here. Further, we note that Reviewer 2 found our focus on specific predictions explored by the specific expression patterns to be a strength of the paper.

> -L210. it seems that authors only focused on genes that were upregulated to fall into 5 categories. What about genes that showed downregulation in expression patterns ?
>

Please see our response to reviewer 1 on this topic.

> -L258: it is rather surprising that majority of transcripts were upregulated at the low temperature, especially since there was no recovery to allow transcriptional machinery to operate. It has repeatedly observed that at low temperature, close to lower thermal limit, transcriptional activity is very weak. For instance, Teets et al. (Physiol. Genomics 44, 764–777, 2012) did not observe any differentially expressed transcript during the cold treatment and Vesala et al. (Insect Mol. Biol. 21, 107–118.2012) only found a few transcripts (3 or 0) modulated by cold in D. montana and D. virilis, respectively. Please comment.
>

This is an interesting observation that actually strengthens our reactionome approach, as we can quantify not only what is upregulated but the temperature at which that response occurs. We now comment on this in the Discussion (lns 505 - 515):

	In both species, there were 3 -- 4 times more transcripts upregulated at low than high temperatures (Table 1). The degree of upregulation at low temperatures is surprising given previous studies (Vesala et al. 2012; Teets et al. 2012b) that found little transcriptional activity at low temperatures. However, these studies exposed organisms to a few extreme (-10 -- 0°C) temperatures. At these extreme low temperatures we also found few upregulated transcripts (Fig. 3A), whereas the peak of low-temperature transcriptional activation occurred near 10°C (Fig. 3). A potential explanation for this pattern is that increased gene expression functions to support elevated metabolism at moderately cold temperatures, as suggested by the metabolic cold adaptation hypothesis (Addo-Bediako et al. 2002).

> -L362-379: A.picea is more cold-adapted than A. carolinensis. A.picea also shows more transcripts being expressed at cold than A. carolinensis. Hence, this correlated pattern is assumed to support enhanced defense hypothesis, that is, stress-adapted species show stronger response. This discussion is based on global expression patterns of hundreds of genes that show expression variation according to temperature. The metrics used here is the number of responsive transcripts. These global patterns are assumed to determine differential thermal adaptation between species. It is certainly interesting to look a general patterns and numbers of transcripts, but this bulk of transcripts obviously include a number of genes that have nothing to do with thermal response or are indirect consequences of exposure. My point is that general trends cannot explain all. Some genes are functionally essential for thermal response, while other are not, and I wonder whether looking a GOs and pathways would not be more functionally meaningful. I am not fully convinced that higher number of cold-responsive transcripts directly translates higher cold hardiness.
>

We completely agree that general trends cannot completely explain anything, but studies that simply report lists of differentially-expressed genes without strong functional work, as are rife in the field, also have substantial explanatory weaknesses. This project focused on broad patterns, testing alternative mechanistic hypotheses (enhanced response, tolerance, genetic assimilation) that did not depend on the response of any specific GO terms or pathways. However, as suggested, we do look at GO terms that are enriched for the global responses - see the section “Gene set enrichment analysis” in the methods, and subsequent discussion of GO terms (metabolism, transcript regulation, etc) in the Results and Discussion. The results of these analyses, which again are prone to storytelling, are used to support the tests of our specific hypotheses. As as example, we find the result that in response to thermal extremes, the warmer-climate species shuts down metabolism-related genes while the cool-climate species increases metabolism and damage repair responses, including  7 heat-shock proteins, to be quite compelling (third paragraph of Discussion, lns 476 - 482).

> -L388-390: More Hsp transcripts are expressed in A. carolinensis compared to A.picea when exposed to high thermal limit. This is assumed to support tolerance hypothesis. Interpretations depend on whether Hsp expression is viewed as stress reactive or protective effect.
>

Our main line of evidence for the tolerance hypothesis in A. carolinensis comes from requiring higher temperatures to get a transcriptional response than A. picea. That is, A. carolinensis tolerates greater heat before mounting a transcriptional response. This indicates that there must be other mechanisms that support this tolerance. We note that A. carolinensis activates fewer heat shock proteins than A. picea, and suggest other mechanisms (proteome stability, behavioral quiescence) that may underlie A. carolinensis’ increased thermal tolerance.

> -L413-418 : More transcripts are expressed at lower temperature. This pattern is explained by greater genetic and phenotypic variation at the lower temperature extreme.  Does more genetic and phenotypic variation necessary translates into more genes being responsive?
>

See our response above - we have removed this section.

> -L492-496:  It is mentioned that A. Picea responds by increasing metabolism and activating heat shock proteins and other protective molecules. I am not sure that transcriptomics can reach such a level of interpretation. RNAseq shows molecular correlates but can certainly not show direct evidence that proteins abundance and activity or more generally pathways and active molecules are actually altered.

We have changed this statement to (lns 593 - 596):

	A. picea responds by increasing expression of transcripts related to metabolism, heat shock proteins and other protective molecules, whereas A. carolinensis decreases expression of transcripts related to metabolism and likely relies on other mechanisms for thermal tolerance.

> Transcriptomics targets genes expression which is situated upstream of the functional molecules. Moreover, it does not account for downstream regulations, and genes and proteins expression might operate with considerable time lag. The molecular machinery provide ample possibilities for post-transcriptional modifications and regulation, so that a direct relationship between mRNA and protein levels cannot be assumed or expected. In addition, mRNA fluctuations (birth/death) are considered as an important source of stochasticity in gene expression, especially when mRNA abundance is low.

This point is absolutely correct, and we briefly acknowledged it in the line

	Finally, the mapping of changes in gene expression to organismal fitness is far from direct (Feder & Walser 2005), and large differences in patterns of gene expression may have only small effects on organismal fitness.

We have added the reviewer’s caveats (lns 566 - 569):

	In particular, functional protein levels cannot be expected to be fully linked to mRNA abundance due to post-transcriptional modification, regulation, mRNA fluctuations and protein stability (Feder & Walser 2005).
